### 2020-04-04 reports/thesis/chapters/3_compoundPoisson.tex
Corrections

The compound Poisson distribution is studied here because of the compound Poission-like behaviour from the detection of photons \citep{whiting2002signal, elbakri2003efficient, whiting2006properties}. It is defined by defining a latent variable $Y\sim\poisson(\lambda)$ with probability mass function (p.m.f.) $\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!}$ for $y=0,1,2,\dotsc$, where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some independent and identically distributed (i.i.d.) latent random variables with probability density function (p.d.f.) $p_U(u)$ for $i=1,2,3,\dotsc$. Let $X$ be a compound Poisson random variable where
\begin{equation}
  X|Y = \sum_{i=1}^{Y}U_i \ .
  \label{eq:compoundPoisson_X|Y}
\end{equation}
The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation}
  p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0
  \ .
\end{equation}
It should be noted that $X=0$ if and only if $Y=0$ and this happens with probability $\prob(Y=0)=\euler^{-\lambda}$. This implies that $X$ has probability mass at zero and probability density for positive numbers which results in the p.d.f.
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\ 
    \sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
  \end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function.

@@

In an x-ray tube, photons are emitted as a Poisson process \citep{whiting2006properties, cierniak2011x} and each photon has some random energy due to bremsstrahlung and characteristic radiation \citep{sun2012overview}. This is similar to the compound Poisson distribution. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with random energy $U_i$ for $i=1,2,3,\dotsc$ with p.d.f.~$p_U(u)$. The random variables discussed here cover all the latent variables in the compound Poisson.

@@

In an energy integrating detector, the recorded grey value is linear to the energy detected \citep{whiting2006properties}. The grey value $X$ is
\begin{equation}
X|Y = \sum_{i=1}^Y U_i + \epsilon \ .
\end{equation}
This is the compound Poisson distribution with Normal noise added to it. The scale factor $b$ is not included as this can be absorbed into $U$. Using the result that $\expectation\left[X\right]=\expectation\expectation\left[X|Y\right]$ and $\variance\left[X\right] = \variance\expectation\left[X|Y\right] + \expectation\variance\left[X|Y\right]$, the mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = \lambda\expectation\left[U\right]+a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = \lambda \expectation\left[U^2\right]+\kappa
\end{equation}
respectively. Eliminating $\lambda$ obtains
\begin{equation}
\variance\left[X\right] = \dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \expectation\left[X\right] + \kappa - a\dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \ .
\end{equation}
By assuming no beam hardening, $\expectation\left[U\right]$ and $\expectation\left[U^2\right]$ remains constant, thus there is a linear relationship between the variance and expectation of the grey value \citep{yang2009evaluation}. There are other types of detection schemes \citep{whiting2006properties} but it shall be not be considered here.

@@

It can be shown that the compound Poisson-gamma distribution is in the exponential family for fixed $\alpha$ \citep{jorgensen1987exponential}. To show this, the compound Poisson-gamma distribution was parametrised using the following:
\begin{equation}
  p=\frac{2+\alpha}{1+\alpha}
  \ ,
\end{equation}
\begin{equation}
  \mu=\frac{\lambda\alpha}{\beta}
  \ ,
\end{equation}
\begin{equation}
  \phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}}
  \ .
\end{equation}
The parameters $p$, $\mu$ and $\phi$ are called the index, mean and dispersion parameters respectively and take the values of $1<p<2$, $\mu>0$ and $\phi>0$. It can be shown that the p.m.f.~at zero is
\begin{equation}
  \prob(X=0) = \exp
  \left[
      -\frac{\mu^{2-p}}{\phi(2-p)}
  \right]
\end{equation}
and the p.d.f.~for $x>0$ is
\begin{equation}
  p_X(x) = 
  \exp\left[
    \frac{1}{\phi}
    \left(
      x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}
    \right)
  \right]
  \frac{1}{x}
  \sum_{y=1}^{\infty}W_y(x,p,\phi)
\end{equation}
where
\begin{equation}
  W_y = W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
  \ .
\end{equation}
The derivation is shown in Appendix \ref{chapter:appendix_tweedie}. This is in the form of a distribution in the dispersive exponential family.\citep{nelder1972generalized, nelder1972generalized_2, mccullagh1984generalized} for fixed $p$ and $\phi$.

@@

$y_\text{l}$ and $y_\text{u}$ can be chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}$ where $\epsilon$ is some small constant, for example, $\epsilon=\euler^{-37}$ will be better than machine precision in 64 bits \citep{dunn2005series}. To prevent overflow problems, it is advised to calculate each term in the summation in log scale \citep{dunn2005series} by using the equation
\begin{equation}
  \ln\left[
    \sum_{y=y_\text{l}}^{y_\text{u}}W_y
  \right]
  = 
  \ln\left(
    W_{y_\text{max}}
  \right)
  +\ln\sum_{y=y_\text{l}}^{y_\text{u}}
  \exp\left[
    \ln\left(W_y\right)-\ln\left(W_{y_\text{max}}\right)
  \right]
  \ .
\end{equation}

@@

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realisations of $X$ were simulated $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 10\,000 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\dotsc,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then normalised by numerically integrating it using the trapezium rule using the 10\,000 evaluated points.

A Q-Q plot is a plot which compares the empirical quantiles with the theoretical quantiles. Let
\begin{equation}
  F_X(x) = \prob(X\leqslant x)
\end{equation}
and
\begin{equation}
  \widehat{F}_X(x) = \frac{1}{n}\cdot\text{max}
  \left[
    \left(\sum_{i=1}^n\mathbb{I}(x_i\leqslant x)\right)-0.5,0
  \right]
  \ ,
\end{equation}
then a Q-Q plot is a parametric plot which plots $\widehat{F}_X^{-1}(p)$ against $F_X^{-1}(p)$ for $p=\frac{0.5}{n},\frac{1.5}{n},\frac{2.5}{n},\dotsc,\frac{n-0.5}{n}$. If $F_X(x)$ and $\widehat{F}_X(x)$ are similar, a Q-Q plot should be a straight line with gradient 1, intercepting the origin. For the exact method and the saddlepoint approximation, $F_X(x)$ was found numerically by evaluating the p.d.f.~at $10\,000$ equally spaced points from and including the minimum to the maximum of the simulated samples, and then summing the required trapeziums. For the saddlepoint approximation, it does not support zero, thus the numerical integration started at a non-zero value. $\widehat{F}_X^{-1}(p)$ was then calculated by interpolation.

When plotting the probability of obtaining a zero or the p.d.f.~for positive values, confidence intervals were plotted as well. Consider a bin in a histogram, the confidence intervals were obtained by assuming that the frequency in a bin $\sim\poisson(\text{p.d.f.~evaluated at the bin}\times 1\,000 \times \text{bin width})$.

For low $\lambda$ (Figures \ref{fig:compoundPoisson_histogram_1_1_1} to \ref{fig:compoundPoisson_histogram_approx_1_100_1}) there was a chance of simulating zeros. For $\lambda = 1$, the Normal approximation failed to capture the probability mass at zero because the Normal distribution is symmetric and support negative values. The saddlepoint approximation did capture the probability mass at zero as shown by an increase in probability density towards zero. For $\CPoisson(1,100,1)$ in Figure \ref{fig:compoundPoisson_histogram_approx_1_100_1}, the compound Poisson-gamma p.d.f.~contained multiple peaks and the saddlepoint approximation was not flexible enough to capture them. In Figure \ref{fig:compoundPoisson_histogram_1_100_1}, the Q-Q plot for the exact method was quite sensitive at the tails of each peak, perhaps there were a few inaccuracies in the exact evaluation of the p.d.f.~or in the calculation of the numerical calculation of the c.d.f.~and its inversion.

@@

@@

\section{Proposed Model}

@@

By assuming no beam hardening, the distribution of the photon energy does not change with attenuation. As a result, all pixels will detect photons with identical energy distributions, thus, $\alpha$ and $\beta$ are the same for all pixels. Attenuation does affect the photon count, the more material a photon has to attenuate, the lower the number of detectable photons. The amount of attenuation depends on the specific path from the source to a pixel in a detector, so $\lambda_i$ varies from pixel to pixel. Electronic Normal noise $\epsilon_{i,j}\sim\normal(a,\kappa)$ may be added as well.

@@

The model may be simplified by omitting the electronic noise and this can be done by setting $a=0$ and $\kappa=0$. The number of pixels can be set to $N=1$ to simplify the model further. Parameter estimation can be done using gradient methods because the likelihood of the compound Poisson can be evaluated \citep{dunn2005series}. However, the EM algorithm is faster and the model is well set up for it.

@@

The EM algorithm \citep{dempster1977maximum} was proposed to estimate the parameters of a $X\sim\CPoisson(\lambda,\alpha,\beta)$ random variable given a sample of measurements of it $\left\{x_1,x_2,x_3,\dotsc,x_n\right\}$. The use of the EM algorithm for the compound Poisson distribution in XCT have been studied in \cite{elbakri2003statistical, xie2008x, xu2009electronic}.

@@

The EM algorithm \citep{dempster1977maximum} treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotsc, Y_n\right\}$ be realisations of $Y$. Define the joint log-likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.  
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
  \label{eq:compoundPoisson_jointLogLikelihood}
\end{multline}
The estimators were found by optimising the joint log-likelihood. This was done iteratively by estimating the $Y$'s given the $X$'s and parameters (E step), followed by estimating the parameters given the $Y$'s and $X$'s (M step) until some convergence conditions were met.

It will be shown that using some approximations, the E step and M step can be implemented. The Cram\'er-Rao lower bound \citep{rao1945information, cramer1946mathematical} can also be found by using a few approximations. However, simulations showed that these estimators struggle for high $\lambda$.

@@

Let
\begin{equation}
  W_y^{(r)} = y^r W_y \quad \text{for }r=1,2,3,\dotsc
\end{equation}
so that
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
Similarly,
\begin{equation}
  \zeta_i = \variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2
  \ .
\end{equation}
The expectation terms, $y_i$ and $\zeta_i$, are evaluated here so that they can be used in the M step. The evaluation can be done by truncating the sum
\begin{equation}
  \sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. $\ln\left(W^{(r)}_y\right)$ can be expressed as
\begin{equation}
  \ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and taking the derivative with respect to $y$ obtains
\begin{equation}
  \frac{\partial }{\partial y} \ln\left(W_y^{(r)}\right) = \frac{r}{y } + \frac{\partial }{\partial y}\ln\left(W_y\right)
  \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotsc$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
  \frac{\partial }{\partial y} \ln\left(W_y^{(r)}\right)\approx \frac{\partial }{\partial y} \ln\left(W_y\right)
  \ .
\end{equation}
Therefore, the maximum of $W_y^{(r)}$ is located at $y_\text{max}=\dfrac{x^{2-p}}{\phi(2-p)}$ for all $r=0,1,2,\dotsc$. As a result, the same method for evaluating $W_y$ can be used to evaluate $W_y^{(r)}$. For a given $r=0,1,2,\dotsc$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

@@

\subsection{Estimation Error}

The variance of the M step estimators can be obtained by using the Cram\'er-Rao lower bound \citep{rao1945information, cramer1946mathematical} from the Fisher's information matrix \citep{fisher1922on} defined as
\begin{equation}
  \matr{I} = -\expectation\left[
    \nabla_{\lambda,\alpha,\beta}\nabla_{\lambda,\alpha,\beta}\T \ln L(\lambda,\alpha,\beta;X,Y)
  \right]
\end{equation}
where
\begin{equation}
  \nabla_{\lambda, \alpha,\beta}=
  \begin{pmatrix}
    {\partial}/{\partial \lambda}
    \\
    {\partial}/{\partial \alpha}
    \\
    {\partial}/{\partial \beta}
  \end{pmatrix}
  \ .
\end{equation}

Most of the calculations for the elements of the matrix are similar to the derivatives of $T$. For the $\alpha$ component, taking derivatives of the joint log-likelihood (Equation \eqref{eq:compoundPoisson_jointLogLikelihood}) obtains
\begin{align}
\dfrac{\partial\ln L}{\partial\alpha} &= \sum_{i=1}^n\mathbb{I}(x_i>0)
\left[
  Y_i\ln\beta - \psi(Y_i\alpha)Y_i+Y_i\ln x_i
\right]
\\
\dfrac{\partial^2\ln L}{\partial\alpha^2} &= \sum_{i=1}^n\mathbb{I}(x_i>0)
\left[
  -\psi'(Y_i\alpha)Y_i^2
\right] \ .
\end{align}
The expectation of the second-order derivative is
\begin{equation}
-\expectation\left[\dfrac{\partial^2\ln L}{\partial\alpha^2}\right]
=
n\left(1-\euler^{-\lambda}\right)
\expectation\left[
  \psi'(Y\alpha)Y^2
\right]
\end{equation}
where the factor $\left(1-\euler^{-\lambda}\right)$ is the proportion of data where $x_i>0$. $\expectation\left[\psi'(Y\alpha)Y^2\right]$ can be approximated. Let $g(Y)=\psi'(Y\alpha)Y^2$, then
\begin{equation}
\expectation\left[\psi'(Y\alpha)Y^2\right]
\approx
\psi'(\lambda\alpha)\lambda^2
+\dfrac{\lambda}{2}g''(\lambda) \ .
\end{equation}
The derivatives of $g(Y)$ are
\begin{align}
g'(Y)&=\psi''(Y\alpha)Y^2\alpha + 2\psi'(Y\alpha)Y
\\
g''(Y)&=Y^2\alpha^2\psi'''(Y\alpha)+4Y\alpha\psi''(Y\alpha)+2\psi'(Y\alpha)\ .
\end{align}
Therefore,
\begin{multline}
-\expectation\left[\dfrac{\partial^2\ln L}{\partial\alpha^2}\right]\approx n\left(1-\euler^{-\lambda}\right)
\\
\times\left[(\lambda^2+\lambda)\psi'(\lambda\alpha)+2\lambda^2\alpha\psi''(\lambda\alpha)+\dfrac{\lambda^3\alpha^2}{2}\psi'''(\lambda\alpha)\right]
\end{multline}

>>>>

The compound Poisson distribution is studied here because of the compound Poission-like behaviour from the detection of photons \citep{whiting2002signal, elbakri2003efficient, whiting2006properties}. It is defined by defining a latent variable $Y\sim\poisson(\lambda)$ with probability mass function (p.m.f.) $\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!}$ for $y=0,1,2,\ldots$, where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some independent and identically distributed (i.i.d.) latent random variables with probability density function (p.d.f.) $p_U(u)$ for $i=1,2,3,\ldots$. Let $X$ be a compound Poisson random variable where
\begin{equation}
  X|Y = \sum_{i=1}^{Y}U_i \ .
  \label{eq:compoundPoisson_X|Y}
\end{equation}
The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation}
  p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0
  \ .
\end{equation}
It should be noted that $X=0$ if and only if $Y=0$ and this happens with probability $\prob(Y=0)=\euler^{-\lambda}$. This implies that $X$ has probability mass at zero and probability density for positive numbers which results in the p.d.f.
\begin{equation}
  p_X(x) =
  \begin{cases}
    \delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\
    \sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
  \end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function.

@@

In an x-ray tube, photons are emitted as a Poisson process \citep{whiting2006properties, cierniak2011x} and each photon has some random energy due to bremsstrahlung and characteristic radiation \citep{sun2012overview}. This is similar to the compound Poisson distribution. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with random energy $U_i$ for $i=1,2,3,\ldots$ with p.d.f.~$p_U(u)$. The random variables discussed here cover all the latent variables in the compound Poisson.

@@

In an energy integrating detector, the recorded grey value is linear to the energy detected \citep{whiting2006properties}. The grey value $X$ is
\begin{equation}
X|Y = \sum_{i=1}^Y U_i + \epsilon \ .
\end{equation}
This is the compound Poisson distribution with Normal noise added to it. The scale factor $b$ is not included as this can be absorbed into $U$. Using the result that $\expectation\left[X\right]=\expectation\expectation\left[X|Y\right]$ and $\variance\left[X\right] = \variance\expectation\left[X|Y\right] + \expectation\variance\left[X|Y\right]$, the mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = \lambda\expectation\left[U\right]+a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = \lambda \expectation\left[U^2\right]+\kappa
\end{equation}
respectively. Eliminating $\lambda$ obtains
\begin{equation}
\variance\left[X\right] = \dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \expectation\left[X\right] + \kappa - a\dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \ .
\end{equation}
By assuming no beam hardening, $\expectation\left[U\right]$ and $\expectation\left[U^2\right]$ remains constant, thus there is a linear relationship between the variance and expectation of the grey value \citep{yang2009evaluation}. There are other types of detection schemes \citep{whiting2006properties} but it shall not be considered here.

@@

It can be shown that the compound Poisson-gamma distribution is in the exponential family for fixed $\alpha$ \citep{jorgensen1987exponential}. To show this, the compound Poisson-gamma distribution was parametrised using the following:
\begin{equation}
  p=\frac{2+\alpha}{1+\alpha}
  \ ,
\end{equation}
\begin{equation}
  \mu=\frac{\lambda\alpha}{\beta}
  \ ,
\end{equation}
\begin{equation}
  \phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}}
  \ .
\end{equation}
The parameters $p$, $\mu$ and $\phi$ are called the index, mean and dispersion parameters respectively and take the values of $1<p<2$, $\mu>0$ and $\phi>0$. It can be shown that the p.m.f.~at zero is
\begin{equation}
  \prob(X=0) = \exp
  \left[
      -\frac{\mu^{2-p}}{\phi(2-p)}
  \right]
\end{equation}
and the p.d.f.~for $x>0$ is
\begin{equation}
  p_X(x) =
  \exp\left[
    \frac{1}{\phi}
    \left(
      x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}
    \right)
  \right]
  \frac{1}{x}
  \sum_{y=1}^{\infty}W_y(x,p,\phi)
\end{equation}
where
\begin{equation}
  W_y = W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
  \ .
\end{equation}
The derivation is shown in Appendix \ref{chapter:appendix_tweedie}. This is in the form of a distribution in the dispersive exponential family \citep{nelder1972generalized, nelder1972generalized_2, mccullagh1984generalized} for fixed $p$.

@@

The limits, $y_\text{l}$ and $y_\text{u}$, can be chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}$ where $\epsilon$ is some small constant, for example, $\epsilon=\euler^{-37}$ will be better than machine precision in 64 bits \citep{dunn2005series}. To prevent overflow problems, it is advised to calculate each term in the summation in log scale \citep{dunn2005series} by using the equation
\begin{equation}
  \ln\left[
    \sum_{y=y_\text{l}}^{y_\text{u}}W_y
  \right]
  =
  \ln\left(
    W_{y_\text{max}}
  \right)
  +\ln\sum_{y=y_\text{l}}^{y_\text{u}}
  \exp\left[
    \ln\left(W_y\right)-\ln\left(W_{y_\text{max}}\right)
  \right]
  \ .
\end{equation}

@@

The evaluation of the p.d.f.~using the saddlepoint approximate required a bit of caution to avoid over/underflow problems. Suppose realisations of $X$ were simulated $\left\{x_1,x_2,x_3,\ldots,x_n\right\}$. The saddlepoint approximation was computed up to a constant using
\begin{equation}
  p_X(x) \propto
  \exp\left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x)
    -x\beta
    +\left(
    \frac{x\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1) - k
  \right]
\end{equation}
for 10\,000 equally spaced points from and including the minimum non-zero simulated value to the maximum simulated value. $k$ is some constant which was chosen to be
\begin{equation}
  k =
  \max_{i\in\left\{1,2,3,\ldots,n\right\}}
  \left[
    -\frac{\alpha+2}{2(\alpha+1)}
    \ln(x_i)
    -x_i\beta
    +\left(
      \frac{x_i\beta}{\alpha}
    \right)^{\frac{\alpha}{\alpha+1}}\lambda^{\frac{1}{\alpha+1}}(\alpha+1)
  \right]
  \ .
\end{equation}
The density was then normalised by numerically integrating it using the trapezium rule using the 10\,000 evaluated points.

A Q-Q plot is a plot which compares the empirical quantiles with the theoretical quantiles. Let
\begin{equation}
  F_X(x) = \prob(X\leqslant x)
\end{equation}
and
\begin{equation}
  \widehat{F}_X(x) = \frac{1}{n}\cdot\text{max}
  \left[
    \left(\sum_{i=1}^n\mathbb{I}(x_i\leqslant x)\right)-0.5,0
  \right]
  \ ,
\end{equation}
then a Q-Q plot is a parametric plot which plots $\widehat{F}_X^{-1}(p)$ against $F_X^{-1}(p)$ for $p=\frac{0.5}{n},\frac{1.5}{n},\frac{2.5}{n},\ldots,\frac{n-0.5}{n}$. If $F_X(x)$ and $\widehat{F}_X(x)$ are similar, a Q-Q plot should be a straight line with gradient 1, intercepting the origin. For the exact method and the saddlepoint approximation, $F_X(x)$ was found numerically by evaluating the p.d.f.~at $10\,000$ equally spaced points from and including the minimum to the maximum of the simulated samples, and then summing the required trapeziums. For the saddlepoint approximation, it does not support zero, thus the numerical integration started at a non-zero value. $\widehat{F}_X^{-1}(p)$ was then calculated by interpolation.

When plotting the probability of obtaining a zero or the p.d.f.~for positive values, confidence intervals were plotted as well. Consider a bin in a histogram, the confidence intervals were obtained by assuming that the frequency in a bin $\sim\poisson(\text{p.d.f.~evaluated at the bin}\times 1\,000 \times \text{bin width})$. The 68\% confidence interval was chosen because this is a typical choice in physics \citep{cowan1998statistical}.

@@

For low $\lambda$ (Figures \ref{fig:compoundPoisson_histogram_1_1_1} to \ref{fig:compoundPoisson_histogram_approx_1_100_1}) there was a chance of simulating zeros. The exact method has an advantage here because it can evaluate probability mass at zero. The Normal approximation failed to capture the probability mass at zero because the Normal distribution is symmetric and support negative values. The saddlepoint approximation improves on the Normal approximation by capturing the skewness and kurtosis of the distribution \citep{bedrick1992empirical, butler2007saddlepoint}. As a result, it captured the probability mass at zero as shown by an increase in probability density towards zero.

For $\CPoisson(1,100,1)$, the compound Poisson-gamma p.d.f.~contained multiple peaks. Figure \ref{fig:compoundPoisson_histogram_approx_1_100_1} shows that the saddlepoint approximation was not flexible enough to capture them. In Figure \ref{fig:compoundPoisson_histogram_1_100_1}, the Q-Q plot for the exact method was quite sensitive at the tails of each peak, perhaps there were a few inaccuracies in the exact evaluation of the p.d.f.~or in the calculation of the numerical calculation of the c.d.f.~and its inversion.

@@

\section{Proposed Model}
\label{chapter2:proposed_model}

@@

By assuming no beam hardening, the distribution of the photon energy does not change with attenuation. As a result, all pixels will detect photons with identical energy distributions, thus, $\alpha$ and $\beta$ are the same for all pixels. Attenuation does affect the photon count, the more material a photon has to attenuate, the lower the number of detectable photons. The amount of attenuation depends on the specific path from the source to a pixel in a detector, so $\lambda_i$ varies from pixel to pixel. Electronic Normal noise $\epsilon_{i,j}\sim\normal(a,\kappa)$ is added to model noise recorded by the x-ray detector when no x-rays are exposed. This was observed in the black images in Chapter \ref{chapter2}.

@@

The assumption of a gamma-distributed x-ray photon energy and Normal electronic noise was applied to XCT by \cite{xu2009electronic}. The assumptions were verified by investigating the variance of radiographs for different x-ray exposures. More statistically sound methods could use goodness of fit tests or Q-Q plots. The use of the gamma distribution will not be exactly correct because the sources of x-ray photon energy are characteristic and bremsstrahlung radiation which does not behave like a gamma random variable. However, characterisation of the density of the x-ray photon energy is difficult and usually involve more apparatuses and numerical methods \citep{fewell1977photon, stumbo2004direct}.

The model may be simplified by omitting the electronic noise and this can be done by setting $a=0$ and $\kappa=0$. Further simplification can be done by removing spatial variation by setting $\lambda=\lambda_1=\lambda_2=\ldots=\lambda_N$. Parameter estimation can be done using gradient methods because the likelihood of the compound Poisson can be evaluated \citep{dunn2005series}. However, the EM algorithm is faster and the model is well set up for it.

@@

The EM algorithm \citep{dempster1977maximum} was proposed to estimate the parameters of a $X\sim\CPoisson(\lambda,\alpha,\beta)$ random variable given a sample of measurements of it $\left\{x_1,x_2,x_3,\ldots,x_n\right\}$. The use of the EM algorithm for the compound Poisson distribution in XCT have been studied in \cite{elbakri2003statistical, xie2008x, xu2009electronic}.

@@

The EM algorithm \citep{dempster1977maximum} treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\ldots, Y_n\right\}$ be realisations of $Y$. Define the joint log-likelihood to be
\begin{multline*}
  \ln L(\lambda,\alpha,\beta;X,Y)
  =
  \sum_{i=1}^n
  \left[
    \mathbb{I}(x_i=0)
    \ln
    \prob(Y=0)
  \right.
  \\
  \left.
    +
    \mathbb{I}(x_i>0)
    \ln
    \left[
      p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
    \right]
  \right]
\end{multline*}
so that
\begin{multline}
  \ln L(\lambda,\alpha,\beta;X,Y)=
  \sum_{i=1}^n
  \left[
    -\mathbb{I}(x_i=0)
    \lambda
  \right.
  \\
  \left.+
    \mathbb{I}(x_i>0)
    \left(
      Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
    \right.
  \right.
  \\
  \left.
    \left.
      - \lambda + Y_i \ln \lambda - \ln(Y_i!)
    \right)
  \right]
  \ .
  \label{eq:compoundPoisson_jointLogLikelihood}
\end{multline}
The estimators were found by optimising the joint log-likelihood. This was done iteratively by estimating the $Y$'s given the $X$'s and parameters (E step), followed by estimating the parameters given the $Y$'s and $X$'s (M step) until some convergence conditions were met.

It will be shown that using some approximations, the E step and M step can be implemented. However, simulations showed that these estimators struggle for high $\lambda$.

@@

Let
\begin{equation}
  W_y^{(r)} = y^r W_y \quad \text{for }r=1,2,3,\ldots
\end{equation}
so that
\begin{equation}
  y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
Similarly,
\begin{equation}
  \zeta_i = \variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2
  \ .
\end{equation}
The expectation terms, $y_i$ and $\zeta_i$, are evaluated here so that they can be used in the M step. The evaluation can be done by truncating the sum
\begin{equation}
  \sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. This term can be expressed as
\begin{equation}
  \ln W_y^{(r)} =r\ln y+\ln W_y
\end{equation}
and taking the derivative with respect to $y$ obtains
\begin{equation}
  \frac{\partial }{\partial y} \ln W_y^{(r)} = \frac{r}{y } + \frac{\partial }{\partial y}\ln W_y
  \ .
\end{equation}
Keep in mind that $y=1,2,3,\ldots$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
  \frac{\partial }{\partial y} \ln W_y^{(r)}\approx \frac{\partial }{\partial y} \ln W_y
  \ .
\end{equation}
Therefore, the maximum of $W_y^{(r)}$ is located at $y_\text{max}=\dfrac{x^{2-p}}{\phi(2-p)}$ for all $r=0,1,2,\ldots$. As a result, the same method for evaluating $W_y$ can be used to evaluate $W_y^{(r)}$. For a given $r=0,1,2,\ldots$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ were chosen such that $W^{(r)}_{y_\text{l}}$ and $W^{(r)}_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

@@
